Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-11-17 article.tex
Submitted

\documentclass[reprint,amsmath,amssymb,superscriptaddress,showpacs,pre]{revtex4-1}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% \usepackage{hyperref}       % hyperlinks
% \usepackage[numbers]{natbib}
\usepackage{color}      
% \usepackage{url}            % simple URL typesetting
% \usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{enumitem}
\usepackage{csvsimple}
\linenumbers

@@

\author{Pierre Barrat-Charlaix} 
\affiliation{Biozentrum, Universität Basel, Switzerland}
%
\author{John Huddleston}
\affiliation{Molecular Cell Biology, University of Washington, Seattle, WA, USA}
%
\author{Trevor Bedford}
\affiliation{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
%
\author{Richard Neher}
\email{Correspondence to: Richard Neher, Biozentrum, Klingelbergstrasse 70, 4056, Basel, Switzerland. \bf{richard.neher@unibas.ch}}
\affiliation{Biozentrum, Universität Basel, Switzerland}

\begin{abstract}
	Seasonal influenza viruses repeatedly infect humans in part because they rapidly change their antigenic properties and evade host immune responses, necessitating frequent updates of the vaccine composition. 
	Accurate predictions of strains circulating in the future could therefore improve the vaccine match. 
	Here, we studied the predictability of frequency dynamics and fixation of amino acid substitutions.
	Current frequency was strongest predictor of eventual fixation, as expected in neutral evolution. 
    Other properties, such as for example falling in characterized epitopes or high/low \emph{Local Branching Index} (LBI) had little predictive power. 
	Only parallel evolution was found to be predictive of fixation. 
	While the LBI had little power to predict frequency dynamics, is was still successful at picking strains representative of future populations. 
	The latter is due to a tendency of the LBI to be high for consensus-like sequences that are closer to the future than the average sequence. 
	These results are in contrast to simulations of models of adapting populations, where we find the clear signals of predictability. 
	This indicates that the evolution of influenza HA and NA, while driven by strong selection pressure to change, is poorly described by common models of directional selection such as travelling fitness waves.  
\end{abstract}

\maketitle

@@

	Seasonal influenza A viruses (IAV) infect about 10\% of the global population every year, resulting in hundreds of thousands of deaths~\cite{WHOfactsheet,petrova_evolution_2017}. Vaccination is the primary measure to reduce influenza morbidity.
	However, the surface proteins hemagglutinin (HA) and neuraminidase (NA) continuously accumulate mutations at a high rate, leading to frequent antigenic changes \cite{petrova_evolution_2017, Shih6283, bhatt_genomic_2011, Koel976}. 
	While a vaccine targeting a particular strain may be efficient for some time, antigenic drift will sooner or later render it obsolete.
	World Health Organization (WHO) regularly updates influenza vaccine recommendations to best match the circulating strains. 
	Since developing, manufacturing, and distributing the vaccine takes many months, forecasting the evolution of influenza of essential interest to public health \citep{morris2018predictive,klingen_silico_2018}.

	The number of available high quality HA and NA sequences has increased rapidly over the last 20 years~\cite{bogner2006global, shu2017gisaid} and virus evolution and dynamics can be now be tracked at high temporal and spatial resolution \cite{rambaut_genomic_2008}.
	This wealth of data has given rise to an active field of predicting influenza virus evolution ~\cite{morris2018predictive,klingen_silico_2018}. 
	These models predict the future population of influenza viruses by estimating strain fitness or use proxies of fitness.
	\citet{luksza_predictive_2014}, for example, train a fitness model to capture antigenic drift and protein stability on patterns of epitope and non-epitope mutations. 
	Other approaches by \citet{Steinbrueck12123, neher_prediction_2016} use  hemagglutination inhibition (HI) data to determine possible antigenic drift of clades in the genealogy of the HA protein to predict. 
	Finally, \citet{neher_predicting_2014} use branching patterns of HA phylogenies as a proxy for fitness. 
	These branching patterns are summarized by the Local Branching Index (LBI), which was shown to be a proxy of relative fitness in mathematical models of rapidly adapting populations \citep{neher_predicting_2014}.

	The underlying assumption of all these methods is that (i) differences in growth rate between strains can be estimated from sequence or antigenic data and (ii) that these growth rate differences persist for long enough to be predictive of future success. 
	Specific positions in surface proteins are of particular interest in this context.
	The surface proteins are under a strong positive selection and change their amino acid sequence much more rapidly than other IAV proteins or than expected under neutral evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}. 
	Epitope positions, i.e., positions targeted by human antibodies, are expected to change particularly often since virus with altered epitopes can evade existing immune responses \citep{Koel976,wolf_long_2006,Shih6283}.
    It therefore seems plausible that mutations at these positions have a tendency to increase fitness and a higher probability of fixation \citep{strelkowa_clonal_2012}. 
    But one has to be careful to account for the fact that these positions are often ascertained post-hoc \citep{Shih6283} and human immune responses are diverse with substantial inter-individual variation
    \citep{lee_mapping_2019}.

	In this work, we use A/H3N2 HA and NA sequences from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions. 
	This quasi-neutrality is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. 
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of A/H3N2 influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  \\
There are different ways to make this question more specific. First, one can try to quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
A simpler, more qualitative question, is to ask whether $X$ will fix in the population, will disappear, or whether the site will stay polymorphic.

In this work, we use amino-acid sequences of the HA and NA genes of A/H3N2 influenza since the year 2000 available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors). 
This amounts to $44\,976$ HA and $36\,300$ NA sequences, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the A/H3N2 population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).

The central quantities that are derived from this data are \emph{frequency trajectories} of amino acids at each position in the sequences. If an amino acid $X_i$ is found at position $i$ at a frequency between 5\% and 95\% in the population of a given time bin $t$, then the population is considered polymorphic at position $i$ and at time $t$. This polymorphism is characterized by the frequency of $X_i$, $f_{X_i}(t)$, and also by frequencies of other amino acids at $i$. The series of values $f_{X_i}(t)$ for contiguous time bins constitutes the frequency trajectory of $X_i$. A trajectory is terminated if the corresponding frequency is measured above 95\% (resp.~below 5\%) for two time bins in a row, in which case amino acid $X_i$ is considered as \emph{fixed} (resp.~\emph{absent}) in the population. Otherwise, the trajectory is considered \emph{active}. Examples of trajectories can be seen in figure~\sref{fig:sample_trajectories} of the Supplement.

%Frequency trajectories allow a simple formulation of the above-stated questions. The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$. On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes. 
In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. 
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency, should enrich for mutations that are positively selected and on their way to fixation.

@@

To answer this question retrospectively, we use all frequency trajectories extracted from A/H3N2 HA sequences that satisfy these conditions for a given $f_0$.  The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise. 
Therefore, we consider trajectories in an interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$.

For $f_0=0.3$, we find 69 such trajectories, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. 
Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. 
This is due to the fact that some trajectories ``skipped" the interval $f_0$ in question on their initial rise due to sparse sampling. 
These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$. 
Removing them does not change results significantly.

@@

However, future distributions for $\Delta t >0$ do not seem to behave according to this hypothesis. 
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$.
The fact that this average rose $t<0$ has hence no information for $t>0$ and is due to the conditions by which these trajectories were selected.

Consistent with the average, the frequency distribution of the selected trajectories broadens in time without a significant shift of the mean as time passes. 
After 60 days, the distribution is rather symmetrical around the initial $f_0=0.3$ value, suggesting that the knowledge that the trajectories were rising is lost after two months. 
On a timescale of 60 to 120 days, the only possible prediction is that trajectories are likely to be found in a broad interval around the initial frequency $f_0$. 
After one year the distribution becomes almost flat (excluding mutations that have disappeared or fixed), and the initial peak at $f_0$ is not visible anymore. 
The only information remaining from the initial frequency is the fraction that fixed or was lost (see below). 
This behavior is expected in neutral models of evolution \citep{kimura_diffusion_1964} but incompatible with the notion of classic sweeps which take over the population.

While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is uninformative.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$. }
	\label{fig:trajectories}
\end{figure*}

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the more modest goal of predicting the probability that a mutation fixes in the population. 
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency. 
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
We also note that the fraction of lost trajectories increases sharply at small times, indicating that 40\% of mutations observed above 25\% frequency are lost within one year, while it takes longer to fix a mutation in the whole population.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} shows $P_{fix}(f)$ as a function of $f$ for NA and HA and for influenza A/H3N2 and A/H1N1. 
\discuss{For both these proteins, the probability of fixation of a new mutation at frequency $f$ is very close to $f$ itself, that is $P_{fix}(f)\simeq f$.} 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
Indeed, in a neutrally evolving and structure-less population of constant size $N$ such as in the Wright-Fisher model~\cite{kimura_diffusion_1964}, the probability that any individual ultimately becomes the ancestor of all the future population is $1/N$. 
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation. \discuss{Add smth for A/H1N1}

\discuss{This quasi-neutral dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a higher frequency than the one at which they are measured. 
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.}

To investigate further, we try to find categories of mutations that deviate from the neutral expectation of panel \textbf{C} of figure~\ref{fig:fix_loss_active}. 
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution. 
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. 
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
Both groups are again consistent with neutral evolution, suggesting that LBI carries very little information on the probability of fixation of a mutation.

>>>>

\documentclass[reprint,amsmath,amssymb,superscriptaddress,showpacs,rmp]{revtex4-1}

@@

\usepackage[utf8]{inputenc} % allow utf-8 input
% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% \usepackage{hyperref}       % hyperlinks
% \usepackage[numbers]{natbib}
\usepackage{color}      
% \usepackage{url}            % simple URL typesetting
% \usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{lineno}
\usepackage{enumitem}
\usepackage{csvsimple}
% \linenumbers

@@

\author{Pierre Barrat-Charlaix} 
\affiliation{Biozentrum, Universität Basel, Switzerland}
\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland}
%
\author{John Huddleston}
\affiliation{Molecular Cell Biology, University of Washington, Seattle, WA, USA}
%
\author{Trevor Bedford}
\affiliation{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
%
\author{Richard A.~Neher}
\email{Correspondence to: Richard Neher, Biozentrum, Klingelbergstrasse 70, 4056, Basel, Switzerland. \bf{richard.neher@unibas.ch}}
\affiliation{Biozentrum, Universität Basel, Switzerland}
\affiliation{Swiss Institute of Bioinformatics, Basel, Switzerland}

\begin{abstract}
	Seasonal influenza viruses repeatedly infect humans in part because they rapidly change their antigenic properties and evade host immune responses, necessitating frequent updates of the vaccine composition. 
	Accurate predictions of strains circulating in the future could therefore improve the vaccine match. 
	Here, we studied the predictability of frequency dynamics and fixation of amino acid substitutions.
	Current frequency was the strongest predictor of eventual fixation, as expected in neutral evolution. 
    Other properties, such as occurrence in previously characterized epitopes or high \emph{Local Branching Index} (LBI) had little predictive power. 
	Parallel evolution was found to be moderately predictive of fixation. 
	While the LBI had little power to predict frequency dynamics, it was still successful at picking strains representative of future populations. 
	The latter is due to a tendency of the LBI to be high for consensus-like sequences that are closer to the future than the average sequence. 
	Simulations of models of adapting populations, in contrast, show clear signals of predictability. 
	This indicates that the evolution of influenza HA and NA, while driven by strong selection pressure to change, is poorly described by common models of directional selection such as travelling fitness waves.  
\end{abstract}
\date{\today}
\maketitle

@@

	Seasonal influenza A viruses (IAV) infect about 10\% of the global population every year, resulting in hundreds of thousands of deaths~\cite{WHOfactsheet,petrova_evolution_2017}. Vaccination is the primary measure to reduce influenza morbidity.
	However, the surface proteins hemagglutinin (HA) and neuraminidase (NA) continuously accumulate mutations at a high rate, leading to frequent antigenic changes \cite{petrova_evolution_2017, Shih6283, bhatt_genomic_2011, Koel976}. 
	While a vaccine targeting a particular strain may be efficient for some time, antigenic drift will sooner or later render it obsolete.
	The World Health Organization (WHO) regularly updates influenza vaccine recommendations to best match the circulating strains. 
	Since developing, manufacturing, and distributing the vaccine takes many months, forecasting the evolution of influenza is of essential interest to public health \citep{morris2018predictive,klingen_silico_2018}.

	The number of available high quality HA and NA sequences has increased rapidly over the last 20 years~\cite{bogner2006global, shu2017gisaid} and virus evolution and dynamics can be now be tracked at high temporal and spatial resolution \cite{rambaut_genomic_2008}.
	This wealth of data has given rise to an active field of predicting influenza virus evolution ~\cite{morris2018predictive,klingen_silico_2018}. 
	These models predict the future population of influenza viruses by estimating strain fitness or proxies of fitness.
	\citet{luksza_predictive_2014}, for example, train a fitness model to capture antigenic drift and protein stability on patterns of epitope and non-epitope mutations. 
	Other approaches by \citet{Steinbrueck12123, neher_prediction_2016} predict fitness by using  hemagglutination inhibition (HI) data to determine possible antigenic drift of clades in the genealogy of the HA protein. 
	Finally, \citet{neher_predicting_2014} use branching patterns of HA phylogenies as a proxy for fitness. 
	These branching patterns are summarized by the Local Branching Index (LBI), which was shown to be a proxy of relative fitness in mathematical models of rapidly adapting populations \citep{neher_predicting_2014}.

	The underlying assumption of all these methods is that (i) differences in growth rate between strains can be estimated from sequence or antigenic data and (ii) that these growth rate differences persist for long enough to be predictive of future success. 
	Specific positions in surface proteins are of particular interest in this context.
	The surface proteins are under a strong positive selection and change their amino acid sequence much more rapidly than other IAV proteins or than expected under neutral evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}. 
	Epitope positions, i.e., positions targeted by human antibodies, are expected to change particularly often since viruses with altered epitopes can evade existing immune responses \citep{Koel976,wolf_long_2006,Shih6283}.
    It therefore seems plausible that mutations at these positions have a tendency to increase fitness and a higher probability of fixation \citep{strelkowa_clonal_2012}. 
    But one has to be careful to account for the fact that these positions are often ascertained post-hoc \citep{Shih6283} and human immune responses are diverse with substantial inter-individual variation
    \citep{lee_mapping_2019}.

	In this work, we use HA and NA sequences of A/H3N2 and A/H1N1pdm influenza from year 2000 to 2019 to perform a retrospective analysis of frequency trajectories of amino acid mutations. 
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not. 
	To our surprise, we find that the predictability of these trajectories is very limited: The probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions. 
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability. 
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}. 
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?  
The trajectory of a mutation will depend on its own effect on fitness, the contribution of the genetic background on the same segment, and the effect of the remaining seven segments. 
Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to the mutation.

First, we ask whether we can quantitatively predict the frequency of $X$ at future times $f(t)$. In other words, having observed a mutation at frequencies $(f_1, f_2,\ldots,f_n)$ at dates $(t_1,t_2,\ldots,t_n)$, what can we say about its frequency at future dates $(t_{n+1}, t_{n+2},\ldots)$?
A simpler, more qualitative question, is to ask whether $X$ will fix in the population, will disappear, or whether the site will stay polymorphic.

We use amino-acid sequences of the HA and NA genes of A/H3N2 (since the year 2000) and A/H1N1pdm (since the year 2009) influenza available in GISAID \citep{shu2017gisaid} (see supplementary materials for an acknowledgment of all data contributors). 
This amounts to $44\,976$ HA and $36\,300$ NA sequences for A/H3N2 and $45\,350$ HA and $40\,412$ NA sequences for A/H1N1, with a minimum of 100 per year. These sequences are binned in non-overlapping intervals of one month. Each single-month time bin and the sequences that it contains represent a (noisy) snapshot of the influenza population at a given date. The number of sequences per time bin varies strongly both with year and according to the season, with earlier time bins containing around 10 sequences while more recent bins contain several hundreds (see figures~\sref{fig:nseq_per_year} and \sref{fig:nseq_per_month} in SM for details).

The central quantities that we derived from this data are \emph{frequency trajectories} of amino acids at each position in the sequences. If an amino acid $X_i$ is found at position $i$ at a frequency between 5\% and 95\% in the population of a given time bin $t$, then the population is considered polymorphic at position $i$ and at time $t$. This polymorphism is characterized by the frequency of $X_i$, $f_{X_i}(t)$, and also by frequencies of other amino acids at $i$. The series of values $f_{X_i}(t)$ for contiguous time bins constitutes the frequency trajectory of $X_i$. A trajectory is terminated if the corresponding frequency is measured above 95\% (resp.~below 5\%) for two time bins in a row, in which case amino acid $X_i$ is considered as \emph{fixed} (resp.~\emph{absent}) in the population. Otherwise, the trajectory is considered \emph{active}. Examples of trajectories can be seen in figure~\sref{fig:sample_trajectories} of the Supplement.

%Frequency trajectories allow a simple formulation of the above-stated questions. The short term prediction amounts to predicting the curve $f(t)$ starting from a given date $t_0$. On the other hand, the long term prediction amounts to predicting how long a trajectory will remain active, and whether it will fix or disappear when it finishes. 
In the rest of this work, we will focus on frequency trajectories that are starting at a zero (low) frequency, \emph{i.e.} $f(t=0)=0$. 
These represent new amino acid variants which were absent in the population at the time bin when the trajectory started and are currently rising in the population (see Methods).
Such distinction in novel and ancestral variants is necessary to meaningfully interrogate predictability.
Each rising trajectory of a new mutation implies the existence of another decreasing one at the same position, since frequencies of all amino acids at a given position must sum to one. 
If novel variants arise by selection, we expect to see a stronger signal of selection after conditioning on these novel variants. 
In classic models of population genetics, strongly advantageous variants undergo rapid selective \emph{sweeps}, i.e., the rapid rise and fixation. 
The sweep of a mutation can be due to its own fitness effect, to the genetic background or to the effect of the seven other segments. 
By considering the ensemble of novel variants that are rising in frequency, we effectively average over backgrounds, obtaining a set of mutations that we expect to be beneficial on average.
If such sweeps are common in the evolution of HA and NA, the restriction to trajectories that start at low frequency should thus enrich for mutations that are positively selected and on their way to fixation.

@@

To answer this question retrospectively, we use all frequency trajectories extracted from HA and NA sequences that satisfy these conditions for a given $f_0$.  The number of trajectories is limited and the frequency estimates themselves are based on a finite sample and are hence imprecise. 
Therefore, we consider trajectories in an interval $[f_0-\delta f, f_0+\delta f]$ with $\delta f = 0.05$.

For $f_0=0.3$, we found $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. The same analysis was performed for A/H1N1pdm, with the $89$ found trajectories displayed in figure \sref{fig:trajectories_h1n1}.
Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time. 
This is due to the fact that some trajectories ``skipped" the interval $f_0$ in question on their initial rise due to sparse sampling. 
These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$. 
Removing them does not change results significantly.

@@

However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps. 
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories. 
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.

Consistent with the average, the frequency distribution of the selected trajectories broadens in time without a significant shift of the mean as time passes. 
After 60 days, the distribution is rather symmetrical around the initial $f_0=0.3$ value, suggesting that the knowledge that the trajectories were rising is lost after two months. 
On a timescale of 60 to 120 days, the only possible prediction is that trajectories are likely to be found in a broad interval around the initial frequency $f_0$. 
After one year the distribution becomes almost flat (excluding mutations that have disappeared or fixed), and the initial peak at $f_0$ is not visible anymore. 
The only information remaining from the initial frequency is the fraction that fixed or was lost (see below). 
This behavior is expected in neutral models of evolution \citep{kimura_diffusion_1964} but incompatible with a dynamic dominated by sweeps taking over the population.

While this observation does not rule out that signatures exist that predict future frequency dynamics, past dynamics alone is weakly informative.

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$. \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$. }
	\label{fig:trajectories}
\end{figure*}

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the long-term goal of predicting the probability that a mutation fixes in the population. 
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen. 
Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency. 
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}. 
We also note that the fraction of lost trajectories increases sharply at small times with 40\% of mutations observed above 25\% frequency being lost within one year for A/H3N2, while it takes longer to fix a mutation in the whole population.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen. 
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time. 
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA. 
For both proteins, the probability of fixation of a new mutation at frequency $f$ is close to $f$ itself, that is $P_{fix}(f)\simeq f$. 
This result is exactly what is expected in a population evolving in the \emph{absence} of selection. 
% Within the diffusion approximation of the Wright-Fisher model~\cite{kimura_diffusion_1964}, the fixation probability of a mutation with fitness effect $s$ at frequency $f$ is
% \begin{equation}
%     P_{fix}(f) = \frac{1-e^{-2sfN}}{1-e^{-2sN}} \approx
%     \begin{cases}
%     1-e^{-2sfN} & sfN \gg 1 \\
%     f & |sN| \ll 1 \\
%     e^{-2s(1-f)N} & -sN \gg 1  
%     \end{cases}
% \end{equation}
% The result for the neutral case is intuitive:
A mutation or trait appearing at frequency $f$ is shared by $f\cdot N$ individuals, and the probability for one of them to become the ancestor of all the future population is $f\cdot N/N=f$. 
Thus, the probability of this mutation or trait to fix in the population is equal to its current frequency, a case which we will refer to as the neutral expectation. 
Panel \textbf{C} of figure~\ref{fig:fix_loss_active} indicates that mutations in the surface proteins of A/H3N2 influenza are in good agreement with the neutral expectation, while those in A/H1N1pdm show only small deviations from it. 
In both cases, the probability of fixation seems to be mainly dictated by the current frequency $f$ at which the mutation is observed.

This dynamics is in apparent contradiction with evidence that influenza surface proteins are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are measured. 
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

Next, we searched for features of mutations that allow prediction of fixation beyond frequency by dividing frequencies into categories that deviate from the diagonal in panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active}. 
We first turn to the \emph{Local Branching Index} (LBI), a quantity calculated for each node in a phylogenetic tree that indicates how dense the branching of the tree is around that node. 
LBI has previously been successfully used as a predictor of the future population of influenza~\cite{neher_predicting_2014}, and was shown to be a proxy for fitness of leaves or ancestral nodes in mathematical models of evolution. 
Here, we define the LBI of a mutation at date $t$ as the average LBI of strains that carry this mutation and that were sampled in the time bin corresponding to $t$. 
Panel A of figure~\ref{fig:pfix_stratified} shows fixation probability for HA mutations with LBI in the top or bottom half of the distribution. 
Both groups have identical probability of fixation, suggesting that LBI carries very little information on the probability of fixation of a mutation.

Next, we focused on previously reported antigenic sites in the A/H3N2 HA protein, referred to as \emph{epitope} positions. 
Mutations at these position might mediate immune escape and are therefore likely under strong selection and show sweep-like behavior. 
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}. 
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes. 
Only mutations at the 7 epitope sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance. 
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
One should also note that many of these positions were determined post-hoc and might be enriched for positions that experienced rapid substitutions before the publication of the respective studies.
